Msprime 1.0: an efficient and extensible coalescent simulation framework

### 2021-07-23 paper.tex
paragraphs about runtime

\newcommand{\jkcomment}[1]{\textcolor{red}{#1}}
\newcommand{\plrcomment}[1]{\textcolor{purple}{#1}}

@@

\begin{figure}
\begin{center}
\includegraphics[width=6cm]{figures/example_tree_sequence}
\end{center}
\caption{\label{fig-ts-example} Example tree sequence.
An example tree sequence describing genealogies and genotypes
for four samples at ten sites.
}
\end{figure}

@@

\begin{figure}
    \includegraphics{figures/unmutated_tree}
    \includegraphics{figures/mutated_tree}
    \caption{
        \label{fig-mutated-trees}
        Visualization of the separation between ancestry and mutation simulation:
        \texttt{sim\_ancestry} produces ancestry
        (\textbf{left;} shown are two trees along a 1Kb chunk of genome
        relating 200 samples),
        and \texttt{sim\_mutations} adds mutations
        (\textbf{right;} fourteen nucleotide mutations have been added).
        Graphics produced with \tskit's \texttt{draw\_svg} method.
        \plrcomment{TODO: fix up the axis labels, if we like this figure?}
    }
\end{figure}

@@

In human-like parameter regimes, \msprime's implementation of the
exact coalescent with recombination comprehensively outperforms
all other simulators, including those based on SMC
approximations~\citep{kelleher2016efficient}. However, it is important
to note that although the implementation of Hudson's algorithm is
very efficient, it is still quadratic in the scaled recombination rate
$\rho$. This is because Hudson's algorithms tracks recombinations not only
in segments ancestral to the sample, but also between ancestral segments.
Estimates of the number of such ``trapped'' recombinations grow quadratically
with genome length \citep[Eq.~5.10]{hein2004gene}.
Because they do not track such recombinations, efficient SMC implementations
will be faster than \msprime\ for sufficiently high recombination rate,
genome length, or $N_e$. %Until we decide to approximate the recombination process
%between distant ancestral segment. Hint hint.

\jkcomment{We want to round out this discussion here by giving some
sort of rough argument for why Hudson's algorithm is quadratic, and discussing
to Fig.~\ref{fig-recombination-perf} to show roughly what we can expect
to simulate in how long. See issue 28.}

@@

\begin{figure}
\begin{center}
\includegraphics[width=7cm]{figures/arg}
\end{center}
\caption{\label{fig-arg} A figure illustrating the relative size of
the full ARG representation compared the minimal tree sequence.
\jkcomment{This figure is quite preliminary and we need to improve it.}}
\end{figure}

>>>>

\newcommand{\jkcomment}[1]{\textcolor{red}{#1}}
% @plr - purple doesn't exist in my Latex
\newcommand{\plrcomment}[1]{\textcolor{blue}{#1}}

@@

\begin{figure}
\begin{center}
\includegraphics[width=6cm]{illustrations/example_tree_sequence}
\end{center}
\caption{\label{fig-ts-example} Example tree sequence.
An example tree sequence describing genealogies and genotypes
for four samples at ten sites.
}
\end{figure}

@@

\begin{figure}
    \includegraphics{illustrations/unmutated_tree}
    \includegraphics{illustrations/mutated_tree}
    \caption{
        \label{fig-mutated-trees}
        Visualization of the separation between ancestry and mutation simulation:
        \texttt{sim\_ancestry} produces ancestry
        (\textbf{left;} shown are two trees along a 1Kb chunk of genome
        relating 200 samples),
        and \texttt{sim\_mutations} adds mutations
        (\textbf{right;} fourteen nucleotide mutations have been added).
        Graphics produced with \tskit's \texttt{draw\_svg} method.
        \plrcomment{TODO: fix up the axis labels, if we like this figure?}
    }
\end{figure}

@@

In human-like parameter regimes, \msprime's implementation of the
exact coalescent with recombination comprehensively outperforms
all other simulators, including those based on SMC
approximations~\citep{kelleher2016efficient}. However, it is important
to note that although the implementation of Hudson's algorithm is
very efficient, it is still quadratic in the scaled recombination rate
$\rho = 4 N_e L$, where $L$ is the length of the genome in units of recombination distance.
This is because Hudson's algorithms tracks recombinations not only
in segments ancestral to the sample, but also between ancestral segments.
Estimates of the number of such ``trapped'' recombinations grow quadratically
with genome length \citep[Eq.~5.10]{hein2004gene}.
Because they do not track such recombinations, efficient SMC implementations
will be faster than \msprime\ for sufficiently high recombination rate,
genome length, or $N_e$. %Until we decide to approximate the recombination process
%between distant ancestral segment. Hint hint.

% \jkcomment{We want to round out this discussion here by giving some
% sort of rough argument for why Hudson's algorithm is quadratic, and discussing
% to Fig.~\ref{fig-recombination-perf} to show roughly what we can expect
% to simulate in how long. See issue 28.}

Figure~\ref{fig-recombination-perf} gives a rough idea of the runtime of
a given simulation of a constant size population.
However, what about a simulation with \emph{changing} size?
To understand how runtime depends on demography
it helps to consider \emph{why} runtime is quadratic in $\rho$.
At any point in time, \msprime\ must keep track of some number of lineages,
each of which contains some number of chunks of genetic material.
Common ancestor events reduce the number of lineages,
and recombination events increase their number.
However, with long genomes
only a small fraction of the common ancestor events
will involve overlapping segments of ancesty,
and lead to coalescence in the marginal trees.
Such disjoint segments are often far apart (on average, about distance $L/2$),
and so recombine apart again immediately;
it is these large numbers of rapid yet inconsequential events that lead to the quadratic runtime.
\plrcomment{(this is discussed also in the ARG section; check for overlap)}
The maximum number of lineages occurs when
the increase and decrease in numbers of lineages due to
common ancestor and recombination events balance out;
to get an idea of runtime we can estimate when this occurs.
Suppose that the maximal number of lineages is $M$;
at this time the rate of common ancestor events is $M (M-1) / (4 N_e)$
and the total rate of recombination is $M \ell$,
where $\ell$ is the mean length of genome carried by each lineage
(including ``trapped'' non-ancestral material).
It is known that the maximum number of lineages is reached very quickly \citep{nelson2020accounting},
and at this maximum, recombination and coalescence rates are equal,
so many (roughly half) of these lineages have at least two distant segments of ancestral material
separated by a long segment of genome, implying that $\ell \approx L / 4$.
This implies that $M$ is roughly $L N_e$.
Now, the number of linages decreases gradually, on the coalescent time scale -- so, over roughly $2 N_e$ generations.
Since the total rate of events when there are the maximum number of lineages
happens at rate roughly $L^2 N_e / 4$,
then the total number of events is proportional to $(L N_e)^2$ -- i.e., proportional to $\rho^2$,
as desired.
What does this tell us about time-varying population sizes?
The argument above implies that the work is spread out relatively evenly on the coalescent time scale:
in other words,
more events happen in the time period from 0 to $N_e/2$ generations
than from $N_e/2$ to $N_e$ generations,
but the runtime is not dominated but the more recent time period.

Putting this together:
suppose that population size today is $N_1$,
while longer ago than $T$ generations ago it was $N_2$:
does the runtime depend more on $4 N_1 L$ or $4 N_2 L$?
The answer depends on how $T$ commpares to $N_1$: if $T/N_1$ is large,
then runtime will be similar to a population of size $N_1$;
while if $T/N_1$ is small, it will be similar to a population of size $N_2$.
For instance, in many agricultural species $N_1 \propto 100$, while $N_2 \propto 10^5$,
and the runtime will depend critically on $T$ -- in other words,
simulation will be quick in a species with a strong domestication bottleneck,
and slow otherwise.

@@

\begin{figure}

\begin{center}
    \begin{tikzpicture}[scale=0.45]

    \draw (-2,0) -- (-2,3) -- (0, 3);
    \draw (1,0) -- (1,1);
    \draw (0,1) -- (2,1) -- (2,2) -- (4,2) -- (4,0);
    \draw (3,2) -- (3,5);
    \draw (0,1) -- (0,3);
    \draw (-1, 3) -- (-1, 5) -- (3,5);

    \draw[fill] (-2,0) circle [radius=0.05];
    \draw[fill] (-1,3) circle [radius=0.05];
    \draw[fill] (1,0) circle [radius=0.05];
    \draw[fill] (4,0) circle [radius=0.05];
    \draw[fill] (1,1) circle [radius=0.05];
    \draw[fill] (3,2) circle [radius=0.05];
    \draw[fill] (1,5) circle [radius=0.05];

    \node [below] at (-2,0) {0};
    \node [below] at (1,0) {1};
    \node [below] at (4,0) {2};
    \node [above] at (1,1) {3};
    \node [below right] at (1,1) {$x = 0.3$};
    \node [above right] at (3,2) {4};
    \node [above left] at (-1,3) {5};
    \node [above] at (1,5) {6};

    \end{tikzpicture}
% \begin{center}
% $\equiv$
% \end{center}
% \end{column}
% \begin{column}{5cm}
\includegraphics[width=3.8cm]{illustrations/arg-ts.pdf}
\includegraphics[width=3.8cm]{illustrations/arg-ts-simplified.pdf}
\includegraphics[width=3.8cm]{figures/arg.pdf}

\end{center}
\caption{\label{fig-arg} (A) An simple ARG in which a recombination
occurs at position 0.3; (B) the equivalent topology depicted as a tree
sequence, including the recombination node; (C) the same tree sequence
topology ``simplified'' down to its minimal tree sequence representation;
(D) the fraction of nodes in a tree sequence that are concerned with
extra ARG information in a tree sequences returned using \msprime's
\texttt{record\_full\_arg} mode}
\end{figure}
